---
title: "Step 2 - Classify Pitchers by Roles"
author: "Ben Scartz"
date: "2024-01-23"
output: html_document
---




### Roles:

* Innings
  + opener/starter
  + Middle
  + Late
  + Closer
  
* Length
  + Start
  + Long
  + 1 inning/minimum
  
* Leverage
  + High
  + None
  + Low

* Matchup
  + R/L split
  + Best hitters
  
* Creative?
  (exclude for now)
  

```{r include=FALSE}
library(tidyverse)
```


```{r}
# Load data and libraries
#library(tidyverse)
statcast <- read_csv('savant_pitch_level.csv')
fangraphs_table <- read.csv('fangraphs_cleaned.csv')
```
  

###############################################################################
########################## Innings ############################################
###############################################################################

* Opener / starter

  + >75% of appearances begin in the 1st inning
  
* Middle

  + >75% of appearances begin in the 2nd-6th innings

* Late

  + >75% of appearances begin in the 7th-9th innings
  
* Closer 

  + >90% of appearances begin in the 9th inning
  



```{r}
#create a dataframe called inning_roles

innings_roles <- data.frame()

# Extract unique combinations of pitchers and game years from the statcast dataset

pitchers_df <- unique(statcast[c('pitcher', 'game_year')])

# Loop through each row in the pitcher_df dataframe

for(row in 1:nrow(pitchers_df)){

# Extract pitcher names and game year for each row  
  
pitcher_name <- pitchers_df$pitcher[row]
game_year <- pitchers_df$game_year[row]

# Filter the statcast dataset for entries matching the current pitcher and year
  
player_data <- filter(statcast, pitcher == pitcher_name, game_year == game_year)

# Calculate the first inning pitched in each game and classify innings

table <- player_data %>%
  group_by(game_pk) %>%
  summarize(inning = min(inning)) %>%
  group_by(inning) %>%
  summarize(apps = n(),
            total = nrow(.),
            percentage = n() / nrow(.))

# Assign roles based on the inning number

table <- table %>%
  mutate(class = case_when(inning == 1 ~ 'start',
                                  inning >= 2 &
                                    inning <= 7 ~ 'middle',
                                  inning >= 8  ~ 'late'))

# Aggregate percentages by class to determine the pitcher's role

classes <- table %>%
  group_by(class) %>%
  summarize(percentage = sum(percentage))

# Calculate the percentage for each role, defaulting to 0 if not present

  start_percentage <- ifelse('start' %in% classes$class, classes$percentage[classes$class == 'start'], 0)
  middle_percentage <- ifelse('middle' %in% classes$class, classes$percentage[classes$class == 'middle'], 0)
  late_percentage <- ifelse('late' %in% classes$class, classes$percentage[classes$class == 'late'], 0)
  close_percentage <- ifelse('close' %in% classes$class, classes$percentage[classes$class == 'close'], 0)

# Determine the pitcher's role based on the highest percentage among start, middle, and late
  
new_row <- data.frame(pitcher = pitcher_name,
                      year = game_year,
                          role = if_else(start_percentage > 0.70, 'start',
                                  if_else(middle_percentage > 0.70, 'middle',
                                   if_else(late_percentage > 0.70, 'late', 'none'))))

# Add the new row to the innings_roles dataframe

innings_roles <- rbind(innings_roles, new_row)
}
```

###############################################################################
########################## Length #############################################
###############################################################################

* Short 

  + 1-25 pitches
  
* Medium 

  + 26-40
  
* Long

  + >40 pitches

```{r}

# Initalize an empty dataframe for pitching length

length_roles <- data.frame()

# Extract unique combinations of pitchers and game years from the statcast dataset

pitchers_df <- unique(statcast[c('pitcher', 'game_year')])

# Loop through each row in the pitchers_df dataframe

for(row in 1:nrow(pitchers_df)){
  
# Extract the pitcher's name and game year from each row

pitcher_name <- pitchers_df$pitcher[row]
game_year <- pitchers_df$game_year[row]
  
# Filter the statcast dataset for records matching the current pitcher and year

player_data <- filter(statcast, pitcher == pitcher_name, game_year == game_year)

# Calculate the length of appearance in each game and group by length

table <- player_data %>%
  group_by(game_pk) %>%
  summarize(length = n()) %>%
  group_by(length) %>%
  summarize(total = nrow(.),
            percentage = n() / nrow(.))

# Classify the length of appearances into short, medium, or long

table <- table %>%
  mutate(class = case_when(length <= 25 ~ 'short',
                                  length > 25 &
                                    length <= 40 ~ 'medium',
                                  length > 40  ~ 'long'))

# Aggregate percentages by class to determine the pitcher's length role

classes <- table %>%
  group_by(class) %>%
  summarize(percentage = sum(percentage))

# Calculate the percentage for each length role, defaulting to 0 if not present

  short_percentage <- ifelse('short' %in% classes$class, classes$percentage[classes$class == 'short'], 0)
  medium_percentage <- ifelse('medium' %in% classes$class, classes$percentage[classes$class == 'medium'], 0)
  long_percentage <- ifelse('long' %in% classes$class, classes$percentage[classes$class == 'long'], 0)

# Determine the pitcher's length role based on the highest percentage  
  
new_row <- data.frame(pitcher = pitcher_name,
                      year = game_year,
                          role = if_else(short_percentage > 0.60, 'short',
                                  if_else(medium_percentage > 0.60, 'medium',
                                   if_else(long_percentage > 0.60, 'long', 'none'))))

# Append the new row to the length_roles dataframe

length_roles <- rbind(length_roles, new_row)
}

# Update length_roles dataframe based on specific conditions and join with innings_roles

length_roles <- innings_roles %>%
  left_join(length_roles, by = c('pitcher', 'year')) %>%
  filter(role.x != 'start') %>%
  select(-role.x) %>%
  rename(role = role.y)
  
```

###############################################################################
########################## Leverage ###########################################
###############################################################################
  
* High 

  + gmLI > 1.1704 (Q3)
  
* None
  
  + 1.1704 < gmLI < 0.8788

* Low 

  + gmLI < 0.8788
  
```{r}
#Select relevant columns from fangraphs_table

leverage_table <- fangraphs_table %>%
  select(MLBAMID, NameASCII, Season, gmLI, IP)

# Calculate and display a summary of the gmLI values for pitchers with more than 25 IP

summary(filter(leverage_table, IP > 25)$gmLI)

```

```{r}
# Categorize pitchers into leverage roles based on their gmLI and rename columns for clarity

leverage_roles <- leverage_table %>%
  mutate(role = case_when(gmLI > 1.1704 ~ 'high',
                          gmLI < 1.1704 & 
                            gmLI > 0.8788 ~ 'none',
                          gmLI < 0.8788 ~ 'low')) %>%
  rename(pitcher = MLBAMID,
         name = NameASCII,
         year = Season)

# Update leverage_roles with innings_roles, filter out 'start' roles, and adjust column naming

leverage_roles <- innings_roles %>%
  left_join(leverage_roles, by = c('pitcher', 'year')) %>%
  filter(role.x != 'start') %>%
  select(-role.x) %>%
  rename(role = role.y)

```

###############################################################################
########################## Matchups ###########################################
###############################################################################

* RHH

  + > 66% of batters faced are RHH
  
* LHH

  + > 50% of batters faced are LHH
  
```{r message=FALSE, warning=FALSE}

# Initialize an empty dataframe to store matchup roles for pitchers

matchup_roles <- data.frame()

# Extract unique combinations of pitchers and game years from the statcast dataset

pitchers_df <- unique(statcast[c('pitcher', 'game_year')])

#Loop through each row in the pitcher_df dataframe

for(row in 1:nrow(pitchers_df)){

# Extract the pitcher's name and game year for the current row  
  
pitcher_name <- pitchers_df$pitcher[row]
game_year <- pitchers_df$game_year[row]

# Filter the statcast dataset for records matching the current pitcher and year
  
player_data <- filter(statcast, pitcher == pitcher_name, game_year == game_year)

# Group by game and at-bat number, then summarize the batter's side for each at-bat

table <- player_data %>%
  group_by(game_pk, at_bat_number) %>%
  summarize(batter_side = first(stand)) %>%
  mutate(R = if_else(batter_side == 'R', 1, 0),
         L = if_else(batter_side == 'L', 1, 0))
  
# Calculate the proportion of matchups against RHH and LHH

R <- sum(table$R) / nrow(table)
L <- sum(table$L) / nrow(table)
         

# Assign a role based on the proportion of RHH and LHH faced
         
new_row <- data.frame(pitcher = pitcher_name,
                      year = game_year,
                      role = case_when(R > 0.66 ~ 'RHH',
                                       L > 0.50 ~ 'LHH',
                                       TRUE ~ 'none'))

# Append the new row to the matchup_roles dataframe

matchup_roles <- rbind(matchup_roles, new_row)

}

#Update matchup_roles with innings_roles, filter out 'start' roles, and adjust column naming

matchup_roles <- innings_roles %>%
  left_join(matchup_roles, by = c('pitcher', 'year')) %>%
  filter(role.x != 'start') %>%
  select(-role.x) %>%
  rename(role = role.y)
```


###############################################################################
######################## Combine Roles ########################################
###############################################################################

```{r}
#Rename the role columns in the dataframe

matchup_roles <- rename(matchup_roles, matchup_role = role)
leverage_roles <- rename(leverage_roles, leverage_role = role)
length_roles <- rename(length_roles, length_role = role)
innings_roles <- rename(innings_roles, inning_role = role)

# Merge the roles dataframes on pitcher ID and year to create a comprehensive pitcher_roles dataframe

pitcher_roles <- innings_roles %>%
  left_join(matchup_roles, by = c('pitcher', 'year')) %>%
  left_join(leverage_roles, by = c('pitcher', 'year')) %>%
  left_join(length_roles, by = c('pitcher', 'year')) %>%
  select(pitcher, year, inning_role, length_role, matchup_role, leverage_role)

```


```{r write csv}
write.csv(pitcher_roles, 'pitcher_roles.csv')
```


```{r}
pitcher_roles <- read.csv('pitcher_roles.csv')
```










