---
title: "Step 3"
author: "Jack Arbuckle"
date: "2024-01-26"
output: html_document
---
```{r warning=FALSE}
library(tidyverse)
```

### Outline Characteristics and performance metrics

* Pitch arsenal

  + Usage
  + Movement / velo
  + VAA
  + HAA
  + Stuff
  + # of plus stuff pitches
  
* Performance metrics

  + xwOBA against (include Vs RHH/LHH)
  + BB%
  + K%
  + K/out
  + wFC_sc (wFC/C) etc.
  
 
```{r}
# read in the data

fangraphs_table <- read.csv('fangraphs_cleaned.csv')
statcast <- read.csv('savant_pitch_level.csv')
```


###############################################################################    ################################# Pitch Usages ################################ ###############################################################################

```{r}
#Select the relevant columns for pitch usage

pitch_usage <- fangraphs_table %>%
  select(MLBAMID, Season, FA_pct,SL_pct,CT_pct,CB_pct,CH_pct,SF_pct,KN_pct,XX_pct) %>%
  rename(pitcher = MLBAMID,
         year = Season)
```


###############################################################################    ####################### Movement / velo by pitch ############################## ###############################################################################

```{r}
#Select the relevant columns for pitch characteristics (velo and movement)

pitch_velo_movement<- fangraphs_table %>%
  select(MLBAMID, Season, vFA_sc, vFC_sc,vFS_sc, vFO_sc, vSI_sc,vSL_sc,vCU_sc,vKC_sc,vEP_sc,vCH_sc,vSC_sc,vKN_sc,FA_X_sc,FC_X_sc,FS_X_sc,FO_X_sc,SI_X_sc,SL_X_sc,CU_X_sc,KC_X_sc,EP_X_sc,CH_X_sc,SC_X_sc,KN_X_sc,FA_Z_sc,FC_Z_sc,FS_Z_sc,FO_Z_sc,SI_Z_sc,SL_Z_sc,CU_Z_sc,KC_Z_sc
,EP_Z_sc,CH_Z_sc,SC_Z_sc,KN_Z_sc) %>%
  rename(pitcher = MLBAMID,
         year = Season)

```

###############################################################################    ################################# VAA by pitch ################################ ###############################################################################

```{r}
#Create the Vertical Approach Angle statistic for each pitch
statcast <- statcast %>%
    mutate(
    vy_f = -sqrt(vy0^2 - (2 * ay * (50 - (17/12)))),
    t = (vy_f - vy0) / ay,
    vz_f = vz0 + (az * t),
    VAA = -atan(vz_f / vy_f) * (180 / pi)
  )

# Group data by pitcher, game year, and pitch type, then calculate the mean VAA for each group
VAA_table <- statcast %>%
  group_by(pitcher, game_year, pitch_type) %>%
  summarise(mean_VAA = mean(VAA), .groups = 'drop') %>%  # Calculate mean VAA and drop grouping
  mutate(pitch_type = ifelse(is.na(pitch_type) | pitch_type == "", "NA", pitch_type)) %>%  # Handle missing or empty pitch types
  pivot_wider(id_cols = c("pitcher", "game_year"),  # Pivot the data to a wide format
              names_from = pitch_type,  # Use pitch type as column names
              names_prefix = "VAA_",  # Prefix for new columns to indicate they represent mean VAA values
              values_from = mean_VAA) %>%
  rename(year = game_year)  # Rename 'game_year' column to 'year'
```

###############################################################################    ################################# HAA by Pitch ################################ ###############################################################################

```{r}
#Create the Horizontal Approach Angle statistic for each pitch
statcast <- statcast %>%
  mutate(
    vy_f = -sqrt(vy0^2 - (2 * ay * (50 - (17/12)))),
    t = (vy_f - vy0) / ay,
    vx_f = vx0 + (ax * t),
    HAA = -atan(vx_f / vy_f) * (180 / pi)
  )

# Group data by pitcher, game year, and pitch type, then calculate the mean HAA for each group
HAA_table <- statcast %>%
  group_by(pitcher, game_year, pitch_type) %>%
  summarise(mean_HAA = mean(HAA), .groups = 'drop') %>%  # Calculate mean HAA and drop grouping
  # Handle missing or empty values for pitch_type by assigning them a category "NA"
  mutate(pitch_type = ifelse(is.na(pitch_type) | pitch_type == "", "NA", pitch_type)) %>%
  # Pivot the data to a wide format, making each pitch type's mean HAA a separate column
  pivot_wider(id_cols = c("pitcher", "game_year"), 
              names_from = pitch_type, 
              names_prefix = "HAA_",  # Prefix for new columns to indicate they represent mean HAA values
              values_from = mean_HAA) %>%
  rename(year = game_year)  # Rename 'game_year' column to 'year'
```

###############################################################################    ################################# Stuff+ by pitch ############################# ###############################################################################

```{r}

#select the relevant columns for stuff+
stuff_table <- fangraphs_table %>%
  select( c(2,4,6,277:306))

# Further select the relevant columns for stuff+

overall_grades <- stuff_table %>% 
  select(MLBAMID, Season,Pit_plus_CH, Pit_plus_CU, Pit_plus_FA, Pit_plus_SI, Pit_plus_SL, Pit_plus_KC, Pit_plus_FC, Pit_plus_FS, Pit_plus_FO, Pitching_plus) %>%
    rename(pitcher = MLBAMID,
         year = Season)
```

###############################################################################    ####################### Number of Plus Stuff Pitches ########################## ###############################################################################

```{r}

#remove the MLBAMID and Season column to apply the function below to this new dataframe
overall_grades_1 <- stuff_table %>% 
  select(Pit_plus_CH, Pit_plus_CU, Pit_plus_FA, Pit_plus_SI, Pit_plus_SL, Pit_plus_KC, Pit_plus_FC, Pit_plus_FS, Pit_plus_FO) 
```

```{r}
# Define a function to count the number of pitches with values over 100 for each row

number_of_stuff_pitches <- function(row) {
  count_over_100 <- sum(row > 100, na.rm = T) # Count values over 100, ignore NA values
  
  return(count_over_100)
}

#Apply the function across rows of overall_grades_1 to calculate the number of "stuff" pitches

  overall_grades$NumberofStuffPitches <- apply(overall_grades_1 ,1, number_of_stuff_pitches)

# Select relevant columns from overall_grades, focusing on pitcher, year, and the calculated number of "stuff" pitches  
    
plus_stuff <- overall_grades %>%
  select(pitcher, year, NumberofStuffPitches) %>%
  rename(plus_stuff_pitches = NumberofStuffPitches)
  
```

###############################################################################    ############ xwOBA - against  Also vs RHH and vs LHH ########################## ###############################################################################

```{r}
# estimated_woba_using_speedangle
# woba_denom

# Calculate overall xwOBA for each pitcher and game year, using estimated wOBA where actual wOBA is missing

woba_table <- statcast %>%
  group_by(pitcher, game_year) %>%
  mutate(estimated_woba_using_speedangle = if_else(is.na(estimated_woba_using_speedangle), woba_value, estimated_woba_using_speedangle)) %>%
  summarize(xwoba = sum(estimated_woba_using_speedangle, na.rm = T) /
            sum(woba_denom, na.rm = T))

# Calculate xwOBA specifically against left-handed batters

woba_table_L <- statcast %>%
  filter(stand == 'L') %>%
  group_by(pitcher, game_year) %>%
  mutate(estimated_woba_using_speedangle = if_else(is.na(estimated_woba_using_speedangle), woba_value, estimated_woba_using_speedangle)) %>%
  summarize(xwoba_l = sum(estimated_woba_using_speedangle, na.rm = T) /
            sum(woba_denom, na.rm = T))

# Calculate xwOBA specifically against right-handed batters

woba_table_R <- statcast %>%
  filter(stand == 'R') %>%
  mutate(estimated_woba_using_speedangle = if_else(is.na(estimated_woba_using_speedangle), woba_value, estimated_woba_using_speedangle)) %>%
  group_by(pitcher, game_year) %>%
  summarize(xwoba_r = sum(estimated_woba_using_speedangle, na.rm = T) /
            sum(woba_denom, na.rm = T))

# Merge the overall, left-handed, and right-handed xwOBA values into a single dataframe and rename columns

woba_values <- woba_table %>%
  left_join(woba_table_R, by = c('pitcher', 'game_year')) %>%
  left_join(woba_table_L, by = c('pitcher', 'game_year')) %>%
  rename(year = game_year)
```

###############################################################################    ################################# BB% and K% ################################## ###############################################################################

```{r}
#select the relevant columns for K-percentage and walk percentage
bb_k_rates <- fangraphs_table %>%
  select(MLBAMID, Season, K_pct,BB_pct ) %>%
  rename(pitcher = MLBAMID,
         year = Season)
```

###############################################################################    ########################### K / out  ########################################## ###############################################################################

```{r}
# Select relevant columns from the fangraphs_table and calculate strikeouts per out
strikeouts_per_out <- fangraphs_table %>%
  select(K_per_9, Season, MLBAMID, NameASCII) %>%
  mutate(KsperOut = K_per_9 / 27) %>% #Calculate strikeouts per out
  rename(pitcher = MLBAMID,
         year = Season)
```

###############################################################################    ################################# Pitch Run Value ############################# ###############################################################################

```{r}
# Select the relevant pitch run value columns

pitch_run_value <- fangraphs_table %>%
  select(MLBAMID, Season, wFB_per_c, wSL_per_c, wCT_per_c, wCB_per_c, wCH_per_c, wSF_per_c, wKN_per_c, FA_pct_sc) %>%
  rename(pitcher = MLBAMID,
         year = Season)

```

###############################################################################    ################################# Fastball Endurance  ############################## ###############################################################################


```{r}

fastball_statcast <- statcast %>%
  filter(pitch_type == "FF") %>%
  arrange(game_pk, pitcher, pitch_number_appearance)


fastball_statcast <- fastball_statcast %>%
  group_by(game_year, game_pk, pitcher) %>%
  mutate(rolling_avg_velocity = rollapply(release_speed, width = 3, FUN = mean, fill = NA, align = 'right', partial = TRUE))
head(fastball_statcast)

summary(fastball_statcast$release_speed)

```



```{r}

# Assuming 'fastballs' dataframe is already filtered for fastballs and arranged by pitcherid, game_pk, and pitch_order

fastballs_with_threshold <- fastball_statcast %>%
  group_by(game_year, game_pk, pitcher) %>%
  # Calculate the row number within each game_pk and pitcher group
  mutate(pitch_number_within_game = row_number()) %>%
  # Add a new column with the rolling average velocity for the fifth pitch
  # or use rolling_avg_velocity if there are fewer than five pitches
  mutate(first_pitch_velocity = ifelse(pitch_number_within_game == 3, rolling_avg_velocity, 
                                       ifelse(pitch_number_within_game < 3, rolling_avg_velocity, NA_real_))) %>%
  # Fill NA values with the value from the fifth pitch or the last available pitch for each group
  fill(first_pitch_velocity, .direction = "downup") %>%
  ungroup()



# Calculate the number of pitches until the pitcher matches or exceeds their first pitch velocity
fastballs_matched_threshold <- fastballs_with_threshold %>%
  group_by(game_year, game_pk, pitcher ) %>%
  mutate(matched_first_velocity = rolling_avg_velocity <= first_pitch_velocity - 1.5,
         pitch_count_to_match = cumsum(matched_first_velocity)) %>%
  filter(matched_first_velocity) %>%
  mutate(first_match_pitch = rank(pitch_number_appearance) == 1) %>%
  filter(first_match_pitch) %>%
  ungroup() %>%
  select(game_year, game_pk,pitcher, pitch_number_appearance, rolling_avg_velocity, first_pitch_velocity, pitch_count_to_match)

# This gives you a dataframe with each pitcher's first pitch that matches or exceeds their first pitch velocity in each game

fastball_endurance <- fastballs_matched_threshold %>%
  group_by(game_year, pitcher) %>%
  summarise(avg_pitch_number_endurance = mean(pitch_number_appearance),
            avg_rolling_avg_velocity = mean(rolling_avg_velocity),
            avg_first_pitch_velocity = mean(first_pitch_velocity)) %>%
  rename(year = game_year)


```

###############################################################################    ################################# xwoba By Times Faced  ############################## ###############################################################################



```{r}


# Calculate opposing team's batting average for each pitcher in each game by times through the order
xwoba_by_order <- statcast %>%
  group_by(game_year, game_pk, pitcher, times_faced) %>%
  mutate(estimated_woba_using_speedangle = if_else(is.na(estimated_woba_using_speedangle), woba_value, estimated_woba_using_speedangle)) %>% 
  summarize(xwoba = sum(estimated_woba_using_speedangle, na.rm = T) /
            sum(woba_denom, na.rm = T)) %>%
  ungroup()

# View the resulting dataframe


```

```{r}



xwoba_by_order_wide <- statcast %>%
  group_by(game_year, game_pk, pitcher, times_faced) %>%
  summarize(
    xwoba = sum(if_else(is.na(estimated_woba_using_speedangle), woba_value, estimated_woba_using_speedangle), na.rm = TRUE) /
             sum(woba_denom, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  # Transform the dataset from long to wide
  pivot_wider(
    names_from = times_faced,
    names_prefix = "xwoba_times_faced_",
    values_from = xwoba
  )

xwoba_by_order_summary <- xwoba_by_order_wide %>%
  group_by(game_year, pitcher) %>%
  summarise(xwoba_times_faced_1 = mean(xwoba_times_faced_1, na.rm = TRUE),
            xwoba_times_faced_2 = mean(xwoba_times_faced_2, na.rm = TRUE),
            xwoba_times_faced_3 = mean(xwoba_times_faced_3, na.rm = TRUE),
            xwoba_times_faced_4 = mean(xwoba_times_faced_4, na.rm = TRUE),
            xwoba_times_faced_5 = mean(xwoba_times_faced_5, na.rm = TRUE))

xwoba_by_order_summary <- xwoba_by_order_summary %>%
  mutate(xwoba_faced_diff_1_2 = xwoba_times_faced_2 - xwoba_times_faced_1,
         xwoba_faced_diff_2_3 = xwoba_times_faced_3 - xwoba_times_faced_2) %>%
  rename(year = game_year)

```


###############################################################################    ################################# Pitch Type Percentage by Times Faced  ############################## ###############################################################################

```{r}

# Calculate the pitch type percentages
pitch_type_percentages_wide <- statcast %>%
  group_by(game_year, game_pk, pitcher, times_faced, pitch_type) %>%
  summarize(pitch_count = n(), .groups = 'drop') %>%
  group_by(game_year, game_pk, pitcher, times_faced) %>%
  mutate(total_pitches = sum(pitch_count)) %>%
  ungroup() %>%
  mutate(pitch_type_percentage = pitch_count / total_pitches) %>%
  # Transform the dataset from long to wide for each times_faced and pitch_type
  pivot_wider(
    id_cols = c(game_year, game_pk, pitcher),
    names_from = c(times_faced, pitch_type),
    names_prefix = "pitch_type_percentage_",
    values_from = pitch_type_percentage,
    values_fill = list(pitch_type_percentage = 0)  # Fill missing values with 0
  )

# Summarize the pitch type percentages by mean for each times faced across all games
pitch_type_percentages_summary <- pitch_type_percentages_wide %>%
  group_by(game_year, pitcher) %>%
  summarize(across(starts_with("pitch_type_percentage_"), mean, na.rm = TRUE))

pitch_type_percentages_summary2 <- pitch_type_percentages_summary %>%
  mutate(across(
    .cols = matches("percentage_2"),
    .fns = ~ . - get(str_replace(cur_column(), "2", "1")),
    .names = "diff_{str_replace(.col, 'percentage_2', 'percentage_1_2_diff')}"
  ))


pitch_type_percentages_summary2 <- pitch_type_percentages_summary2 %>%
  mutate(across(
    .cols = matches("percentage_3"),
    .fns = ~ . - get(str_replace(cur_column(), "3", "2")),
    .names = "diff_{str_replace(.col, 'percentage_3', 'percentage_2_3_diff')}"
  )) %>%
  rename(year = game_year)


```








###############################################################################    ################################# Combine Tables ############################## ###############################################################################

```{r}
#combine all the tables 

pitcher_characteristics_metrics <- pitch_usage %>%
  left_join(pitch_velo_movement, by = c('pitcher', 'year')) %>%
  left_join(VAA_table, by = c('pitcher', 'year')) %>%
  left_join(HAA_table, by = c('pitcher', 'year')) %>%
  left_join(overall_grades, by = c('pitcher', 'year')) %>%
  left_join(plus_stuff, by = c('pitcher', 'year')) %>%
  left_join(woba_values, by = c('pitcher', 'year')) %>%
  left_join(bb_k_rates, by = c('pitcher', 'year')) %>%
  left_join(strikeouts_per_out, by = c('pitcher', 'year')) %>%
  left_join(pitch_run_value, by = c('pitcher', 'year')) %>%
  left_join(pitch_type_percentages_summary2, by = c('pitcher', 'year')) %>%
  left_join(xwoba_by_order_summary, by = c('pitcher', 'year')) %>%
  left_join(fastball_endurance, by = c('pitcher', 'year'))

```

```{r}
pitcher_roles <- read.csv("pitcher_roles.csv")
```

```{r}
pitcher_characteristics_metrics <- pitcher_characteristics_metrics %>%
  left_join(pitcher_roles, by = c('pitcher','year'))
```



```{r}

write.csv(pitcher_characteristics_metrics, "full_table.csv")
```

